\lab{GAME -- Asteroids}

\apparatus
\equip{Computer}
\equip{Glowscript -- www.glowscript.org}

\longgoal

The purpose of this activity is to modify an asteroids game so that when a large asteroid explodes into two smaller asteroids, the center of mass continues with the same velocity, as you observed in the experiment of the colliding pucks.

\procedure


\begin{enumerate}

\subsection*{Playing Asteroids}

\item Play the game \emph{Asteroids}.  A link is available from our course web site. Note the actions of the up, down, left, and right arrow keys and how they affect the motion of the spaceship.

\item Copy the template from our course web site, and paste it into a new file in GlowScript.

\item Run the program. Note how the up, down, left, and right keystrokes affect the motion of the spaceship. It's different than in the original Asteroids game. Also, note that hitting an asteroid with a bullet does not make it break up into pieces. (You will add this functionality.)

%\begin{vpythonblock}
%GlowScript 2.1 VPython
%
%#keyboard contols
%def control(event):
%    global thrust
%    if event.type=='keydown':
%        k = event.which
%        if k == 38:
%            thrust=6
%#        if k == 40:
%#            thrust=0
%        if k == 37:
%            spaceship.rotate(angle=-dtheta,axis=vector(0,0,-1))
%        if k == 39:
%            spaceship.rotate(angle=dtheta,axis=vector(0,0,-1))
%        if k==32:
%            bullet=sphere(pos=spaceship.pos+spaceship.axis, radius=0.1, color=color.yellow)
%            bullet.v=bulletspeed*norm(spaceship.axis)+spaceship.v
%            bulletsList.append(bullet)
%    if event.type=='keyup':
%        thrust=0
%
%#bind events to the function named control
%scene.bind('keydown keyup', control) 
%
%#converts an angle in degrees to an angle in radians
%def rad(degrees): 
%    radians=degrees*pi/180
%    return radians
%
%#checks for a collision between two spheres
%def collisionSpheres(sphere1, sphere2):
%    dist=mag(sphere1.pos-sphere2.pos)
%    if(dist<sphere1.radius+sphere2.radius):
%        return True
%    else:
%        return False
%
%#checks for a collision between a cone and a sphere
%def collisionConeSphere(c, s):
%
%    #result is the variable that we will return
%    #default is False
%    result=False
%
%    #check pos of cone
%    if(collisionSphereAndPoint(s,c.pos)):
%        result=True
%    #check tip of cone
%    result=False
%    tip=c.pos+c.axis
%    if(collisionSphereAndPoint(s,tip)):
%        result=True
%    #check edge of radius in x-y plane 1
%    r1=c.radius*cross(vector(0,0,1),norm(c.axis))
%    if(collisionSphereAndPoint(s,r1+c.pos)):
%        result=True
%    #check edge of radius in x-y plane 2
%    r2=-c.radius*cross(vector(0,0,1),norm(c.axis))
%    if(collisionSphereAndPoint(s,r2+c.pos)):
%        result=True
%
%    #return result
%    return result
%
%#determines whether a point is within a sphere or not
%#returns boolean
%def collisionSphereAndPoint(sphereObj, targetVector):
%    dist=mag(sphereObj.pos-targetVector)
%    if(dist<sphereObj.radius):
%        return True
%    else:
%        return False
%
%#create functions for selecting random numbers
%
%#return a random integer between the min and max at an interval delta
%def randrange(min, max, delta):
%    range=max-min
%    rnum=Math.random()
%    x=Math.floor((rnum*range+1)) + min
%    return int((x%5)*5)
%    
%#return a random integer in a range from min to max
%def randint(min,max):
%    range=max-(min-1)
%    rnum=Math.random()
%    x=Math.floor(rnum*range+1)+(min-1)
%    return int(x)
%    
%#flip a coin
%def choice(a,b):
%    rnum=Math.random()
%    if rnum <0.5:
%        return a
%    if rnum>0.5:
%        return b
%
%#creates four asteroids, one on each side of the scene
%def createAsteroids():
%
%    #asteroid comes from the right
%    asteroid=sphere(pos=vector(20,0,0), radius=1, color=color.cyan)
%    asteroid.pos.y=randrange(-20,20,5)
%    asteroid.m=1
%    asteroid.v=vector(0,0,0)
%    asteroid.v.x=-randint(1,5)
%    asteroid.v.y=choice(1,-1)*randint(1,5)
%    asteroidList.append(asteroid)
%
%    #asteroid comes from the left
%    asteroid=sphere(pos=vector(-20,0,0), radius=1, color=color.cyan)
%    asteroid.pos.y=randrange(-20,20,5)
%    asteroid.m=1
%    asteroid.v=vector(0,0,0)
%    asteroid.v.x=randint(1,5)
%    asteroid.v.y=choice(1,-1)*randint(1,5)
%    asteroidList.append(asteroid)
%
%    #asteroid comes from the top
%    asteroid=sphere(pos=vector(0,20,0), radius=1, color=color.cyan)
%    asteroid.pos.x=randrange(-20,20,5)
%    asteroid.m=1
%    asteroid.v=vector(0,0,0)
%    asteroid.v.x=choice(1,-1)*randint(1,5)
%    asteroid.v.y=-randint(1,5)
%    asteroidList.append(asteroid)
%
%    #asteroid comes from the bottom
%    asteroid=sphere(pos=vector(0,-20,0), radius=1, color=color.cyan)
%    asteroid.pos.x=randrange(-20,20,5)
%    asteroid.m=1
%    asteroid.v=vector(0,0,0)
%    asteroid.v.x=choice(1,-1)*randint(1,5)
%    asteroid.v.y=randint(1,5)
%    asteroidList.append(asteroid)
%
%
%#scene size
%scene.range=20
%scene.width=700
%scene.height=700
%
%#create the spaceship as a cone
%spaceship = cone(pos=vec(0,0,0), axis=vec(2,0,0), radius=1, color=color.white)
%fire = cone(pos=vec(0,0,0), axis=-spaceship.axis/2, radius=spaceship.radius/2, color=color.orange)
%
%#initial values for mass, velocity, thrust, and net force
%spaceship.m=1
%spaceship.v=vector(0,0,0)
%thrust=0
%Fnet=vector(0,0,0)
%
%#bullets
%bulletspeed=10
%bulletsList=[]
%
%#angle to rotate
%dtheta=rad(10)
%
%#clock
%t=0
%dt=0.005
%
%#asteroids
%Nleft=0 #counter for number of asteroids left in the scene
%asteroidList=[]
%createAsteroids()
%
%while spaceship.visible==True:
%    rate(200)
%
%    Fnet=thrust*norm(spaceship.axis)
%    spaceship.v=spaceship.v+Fnet/spaceship.m*dt
%    spaceship.pos=spaceship.pos+spaceship.v*dt
%    fire.pos=spaceship.pos
%    fire.axis=-spaceship.axis/2
%
%
%    #check if the spaceship goes off screen and wrap
%    if spaceship.pos.x>20 or spaceship.pos.x<-20:
%        spaceship.pos=spaceship.pos-spaceship.v*dt
%        spaceship.pos.x=-spaceship.pos.x
%    if spaceship.pos.y>20 or spaceship.pos.y<-20:
%        spaceship.pos=spaceship.pos-spaceship.v*dt
%        spaceship.pos.y=-spaceship.pos.y
%
%    #update positions of bullets and check if bullets go off screen
%    for thisbullet in bulletsList:
%        if thisbullet.pos.x>20 or thisbullet.pos.x<-20:
%            thisbullet.visible=False
%        if thisbullet.pos.y>20 or thisbullet.pos.y<-20:
%            thisbullet.visible=False
%        if thisbullet.visible != 0:
%            thisbullet.pos=thisbullet.pos+thisbullet.v*dt
%
%    #update positions of asteroids
%    for thisasteroid in asteroidList:
%        if thisasteroid.visible==True:
%            thisasteroid.pos=thisasteroid.pos+thisasteroid.v*dt
%            #check for collision with spaceship
%            if(collisionConeSphere(spaceship,thisasteroid)):
%                spaceship.visible=False
%                fire.visible=False
%            #wrap at edge of screen
%            if thisasteroid.pos.x>20 or thisasteroid.pos.x<-20:
%                thisasteroid.pos=thisasteroid.pos-thisasteroid.v*dt
%                thisasteroid.pos.x=-thisasteroid.pos.x
%            if thisasteroid.pos.y>20 or thisasteroid.pos.y<-20:
%                thisasteroid.pos=thisasteroid.pos-thisasteroid.v*dt
%                thisasteroid.pos.y=-thisasteroid.pos.y
%            #check for collision with bullets
%            for thisbullet in bulletsList:
%                if(collisionSpheres(thisbullet,thisasteroid)and thisbullet.visible==True):
%                   thisasteroid.visible=False
%                   thisbullet.visible=False
%
%
%    Nleft=0 #have to reset this before counting asteroids
%    for thisasteroid in asteroidList:
%        if thisasteroid.visible:
%            Nleft=Nleft+1
%
%    #create more asteroids if all are gone
%    if Nleft==0:
%        createAsteroids()
%
%    #update fire
%    if thrust==0 or spaceship.visible==False:
%        fire.visible=False
%    else:
%        fire.visible=True
%
%    t=t+dt
%\end{vpythonblock}

\item Answer the following questions.

\begin{enumerate}
	\item What are the three functions used for generating random numbers and what is the purpose of each of these functions?
\vspace{0.5in}
	\item What is the magnitude of the thrust of the engine when it is firing?
\vspace{0.5in}
	\item What line number calculates the net force on the spaceship?
\vspace{0.5in}
	\item What line numbers update the velocity and position of the spaceship?
\vspace{0.5in}
	\item What line numbers update the positions of the asteroids?
\vspace{0.5in}
	\item What line numbers update the positions of the bullets?
\vspace{0.5in}
	\item What line number makes the asteroid disappear when it is hit by a bullet?
\vspace{0.5in}
	\item How many asteroids are created when the function \code{createAsteroids()} is called and where do these asteroids come from?
\vspace{0.5in}
	\item The velocities of the asteroids are randomized. What is the maximum possible velocity of an asteroid? What is the minimum possible velocity of an asteroid?  (Note that the directions are randomized as well. By ``maximum velocity'' I am referring to the maximum absolute value of the x and y components of the velocity vector. By ``minimum velocity'', I am referring to the minimum absolute value of the x and y components of the velocity vector.)
\vspace{0.5in}
	\item If you decrease the mass of the spaceship (i.e. change it from 1 to a smaller number like 0.5), how would it affect the spaceship's motion when the engine is firing? Answer the question, then test your answer to see if it is correct, and then comment on whether your observation matched your prediction.
	
	\smallframe{\ }
	
\end{enumerate}

\subsection*{Adding Asteroid Explsions}

In the last chapter, you learned that the center of mass of a two-body system is constant. Whether it is a collision of two objects or an exploding fireworks shell, during the collision or explosion the center of mass velocity of the system remains constant.

In the classic Asteroids game, when a large asteroid is shot with a bullet, it breaks into two fragments. Since the center of mass of the system must remain constant then

\begin{eqnarray*}
	\vec{v}_{ast} & = & \frac{m_1\vec{v}_1+m_2\vec{v}_2}{m_1+m_2} \\
\end{eqnarray*}

where 1 and 2 refer to the two fragments. Define the total mass of the fragments as $M=m_1+m_2$, which is the mass of the asteroid before the explosion. Then,

\begin{eqnarray*}
	\vec{v}_{ast} & = & \frac{m_1\vec{v}_1+m_2\vec{v}_2}{M} \\
\end{eqnarray*}

In our game, let's assume that the asteroid breaks into two equal mass fragments that are each $1/2$ the total mass of the asteroid. Then, $m_1=m_2=1/2M$. Thus,

\begin{eqnarray*}
	\vec{v}_{ast} & = & \frac{1/2M\vec{v}_1+1/2M\vec{v}_2}{M} \\
	\vec{v}_{ast} & = &  \frac{\vec{v}_1+\vec{v}_2}{2}
\end{eqnarray*}

In other words, since the fragments have equal masses, then the asteroid's velocity is the arithmetic mean of the velocities of the fragments (i.e. the sum of their velocities divided by 2). 

In our game, we will randomly assign the velocity of fragment 1 so that it will shoot off with a random speed in a random direction. Then we will calculate the velocity of fragment 2 so that

\begin{eqnarray*}
	\vec{v}_{2} & = & 2\vec{v}_{ast}-\vec{v}_1\\
\end{eqnarray*}

\item Above the \code{while} loop, find the line where the asteroidList is created. Near this point in the program, create a list called \code{fragmentList}. To do this, type the following line:

\begin{myvpython}
fragmentList=[]
\end{myvpython}

\item Near the top of the program where other functions are defined (e.g. after the \code{createAsteroids()} function is defined), write a function that creates the fragments when an asteroid is hit. Type the following code and replace the commented line with the correct calculation for the velocity of fragment 2.

\begin{myvpython}
def createFragments(asteroid):
    fragment1=sphere(pos=asteroid.pos, radius=0.5, color=color.magenta)
    fragment2=sphere(pos=asteroid.pos, radius=0.5, color=color.magenta)
    fragment1.m=0.5
    fragment2.m=0.5
    fragment1.v=vector(0,0,0)
    fragment1.v.x=choice(1,-1)*randint(1,5)
    fragment1.v.y=choice(1,-1)*randint(1,5)
#    fragment2.v=
    fragmentList.append(fragment1)
    fragmentList.append(fragment2)   
\end{myvpython}

\item In the main \code{while} loop, there is an \code{if} statement that checks for a collision between a bullet and asteroid. The line reads:

\begin{myvpython}
if(collisionSpheres(thisbullet,thisasteroid)and thisbullet.visible==True):
\end{myvpython}

Inside this \code{if} statement, call the \code{createFragments} function using:

\begin{myvpython}
                   createFragments(thisasteroid)
\end{myvpython}

This creates the explosion whenever the bullet hits an asteroid.

\item You will need to update the positions of the fragments (i.e. make them move) inside the main \code{while} loop. It's easiest to copy and paste the section of code for the asteroids and change the variables names to the appropriate names for the fragments. You can also copy and paste this code from our course web site. Here's an example of what it should look like:

\begin{myvpython}
    #update positions of fragments
    for thisfragment in fragmentList:
        if thisfragment.visible==True:
            thisfragment.pos=thisfragment.pos+thisfragment.v*dt
            #check for collision with spaceship
            if(collisionConeSphere(spaceship,thisfragment)):
                spaceship.visible=False
                fire.visible=False
            #wrap at edge of screen
            if thisfragment.pos.x>20 or thisfragment.pos.x<-20:
                thisfragment.pos=thisfragment.pos-thisfragment.v*dt
                thisfragment.pos.x=-thisfragment.pos.x
            if thisfragment.pos.y>20 or thisfragment.pos.y<-20:
                thisfragment.pos=thisfragment.pos-thisfragment.v*dt
                thisfragment.pos.y=-thisfragment.pos.y
            #check for collision with bullets
            for thisbullet in bulletsList:
                if(collisionSpheres(thisbullet,thisfragment)and thisbullet.visible==True):
                   thisfragment.visible=False
                   thisbullet.visible=False
\end{myvpython}

\item The section of code that counts how many asteroids are left also needs to count how many fragments are left. After the \code{for} loop that counts the asteroids that are left, add a second for \code{loop} that counts the remaining fragments.

\begin{myvpython}
    for thisfragment in fragmentList:
        if thisfragment.visible:
            Nleft=Nleft+1
\end{myvpython}

\item That should do it. Run your code and fix any errors.

\item Note that we could have written our code more efficiently. Whenever you find that you are copying and pasting the same code over and over, you might want to consider putting it into a function and calling the function. For example, we have to wrap the motion of the spaceship, asteroids, and fragments. We can create a \code{wrap()} function that takes an argument of the object (like spaceship) and checks to see if it is off screen. If it is off screen, it wraps the position of the object. If you notice ways like this to reduce the number of lines of code, feel free to make those changes.

\end{enumerate}

\pagebreak

\analysis

\begin{description}

\item[C] Complete this exercise and do the following.

\begin{enumerate}
	\item Add a point system that tallies points whenever an asteroid is shot.
	\item Print the total number of points at the end of the game when the spaceship collides with an asteroid.
\end{enumerate}


\item[B] Do everything for {\bf C} and the following.

\begin{enumerate}
	\item Add a keystroke for the ``down'' arrow key that rotates the spaceship 180$\degree$ ($\pi$ radians). For the syntax on how to do this, see the left and right arrow keys.
	\item Place the entire \code{while spaceship.visible==True:} loop inside another \code{while True:} loop so that after the spaceship collides with an asteroid the program will not end, but rather it will begin again. If you select the entire \code{while spaceship.visible==True:} loop and hit the TAB key, then the selected code will automatically indent.
	
	\item You will need to reinitialize your variables for the velocity, thrust, and net force on the spaceship, along with your lists, in order to reset the game. You can copy code like the following and paste it inside the \code{while True:} statement but at the end. In other words, when the inner loops ends (by virtue of the spaceship not being visible), then this code block below will run before the outer loops starts again.
		
\begin{myvpython}    
    scene.waitfor("click")

    spaceship.visible=True
    spaceship.v=vector(0,0,0)
    thrust=0
    Fnet=vector(0,0,0)
    
    #bullets
    for b in bulletsList:
        b.visible=False
    bulletsList=[]
    
    #asteroids
    Nleft=0 #counter for number of asteroids left in the scene
    for ast in asteroidList:
        ast.visible=False
    asteroidList=[]

    createAsteroids()
    
    #fragments
    for f in fragmentList:
        f.visible=False
    fragmentList=[]
  \end{myvpython}

	
\end{enumerate}

\item[A] Do everything for {\bf B} with the following modifications and additions.

\begin{enumerate}
	\item Think of an improvement that will require knowledge of physics. Here are some ideas: 
	\begin{enumerate}
		\item asteroids collide with each other
		\item bullets have mass. Every time a bullet is fired, it exerts a backwards force on the spaceship and causes the spaceship to lose mass.
		\item the asteroid fragments should have have the volume of the asteroids and thus should have the appropriate radius.
		\item the asteroid can break into three fragments instead of two.
		\item the fragments can have different masses thought the sum is equal to the mass of the asteroid
	\end{enumerate}
	\item Talk to Dr. T the improvement you want to make.
	\item Implement your idea.
\end{enumerate}


\end{description}

