
\myappendix{project1}{Project 1 -- Developing your first game with constant velocity motion}

\longgoal

For this project, you will develop a game using objects that move with constant velocity.

\section*{Project Guidelines}

You will develop a game that involves objects that move with constant velocity. It is due at midnight Mar 24. Upload your game and documentation via WebAssign. You may request help from Michael Welter and I. We can answer questions, give you advice, and help you troubleshoot syntax or logic errors. However we will be cautious to not write significant portions of your code.\\

\noindent
There are four categories that the project's grade is based on.

\begin{enumerate}
	\item level of difficulty (i.e. "Is the game more interesting than just a ball bouncing back and forth on the screen?" and "Is it sufficiently different from the one developed in class?" and "Is the game fun to play?").  Note that you should be creative, but don't try writing a game that exceeds your abilities at this point. Choose a project that is (1) doable and (2) interesting.
	\item level of independence.  Using resources is good, but you can't copy another program or have someone else write the code for you. Always cite your references, including people who help you.
	\item completeness (i.e.  "Does the simulation run?" and  "Did the program include objects that move with constant velocity?" and "Does the program work as expected?")
	\item quality of documentation (i.e. "Did you include relevant references?" and "Did others test your game?" and "Did you write in a clear, coherent, organized, and grammatically correct way?")
\end{enumerate}

\noindent
You should:

\begin{enumerate}
	\item write a VPython program that includes objects that move with constant velocity.
	\item test your game by having others play it and by asking them for feedback.
	\item answer the questions below.
\end{enumerate}

\section*{Documentation}

You must write a document that answers the following questions. The document should be complete; it should have correct grammar; and it should be easy to read and understand. Save the document as a pdf file and upload it through WebAssign. Quality writing and organization is expected. Answer the following questions.

\begin{enumerate}
	\item What is the purpose of your game?
	\item What are the rules of your game?
	\item How must the game be played (i.e. keystrokes, etc.)?
	\item Is this game like any other game that you've seen or played?
	\item Who played your game and what did you learn as a result of their feedback?
	\item What resources did you use to help you in writing the game?  If you used web sites, people (such as Dr. T or the S.I.), books, or any other resources, you must reference them.
	\item What did you personally get out of this project?
\end{enumerate}
